Collective excitations in competing phases in two and three dimensions

### 2024-01-08 paper/main.tex
Clean unused data

There are a variety of different features. The difference between the two lattices is mostly quantitative.
First and foremost, in the SC phase, see panel (a), there is a sharp peak located at $\omega=0$ in $\mathcal{A}_\text{Phase}$ and a singularity in $\spectral{Higgs}$ located at $\omega=2\Delta_\text{SC}$.
We identify them with the well-known Leggett and Higgs modes in superconductors.
The former can be found at zero energy because our model describes a neutral superfluid wherein there is no coupling of the phase of the order parameter to the electromagnetic potential.
Therefore, this phase may be chosen arbitrarily, and changing it requires no energy which ultimately allows a Goldstone mode to exist in the first place. \textcolor{red}{Zitieren!}

@@

This analysis provides that the real part of $\greens{Phase}$ behaves as $1/\omega^2$ indicating that the peak in $\spectral{Phase}$ is the derivative of a $\delta$ distribution.
Note, that it is located at $\omega=0$ of a perfectly asymmetrical bosonic spectral function and therefore must have 0 weight.
This is given here, because $\int \delta'(\omega) \mathrm{d}\omega = 0$.

\textcolor{red}{Ich spekuliere mal...}
The singularity in $\spectral{Higgs}$ behaves as $1/\sqrt{\omega - 2 \Delta_\text{SC}}$.
Previous studies on the dynamics of the order parameter found damped oscillations after a quench.
These have the frequency $\Omega = 2 \Delta_\text{SC}$ and fell of as $1/\sqrt{t}$.
An inverse Fourier transform of $\spectral{Higgs}$ yields precisely the same behavior, thereby reinforcing the validity of our results.

@@

The plots for the phase peak and the CDW peak in the SC phase can be found in appendix \autoref{sec:fit_greens_functions}.
The SC peak in the CDW phase behaves similarly to the latter and is therefore not shown explicitly.
We get a good agreement with the power-law behavior near the peaks with the exponents $a=-2$ (phase) and $a=-1$ (CDW), respectively.
Therefore, we conclude that the CDW peak corresponds to a $\delta$ distribution, while the phase peak is its derivative.
Note, that the latter is located at $\omega=0$ of a perfectly asymmetrical bosonic spectral function and therefore must have 0 weight.
This is given here, because $\int \delta'(\omega) \mathrm{d}\omega = 0$.
The peak itself can be interpreted as two $\delta$ distributions at $0^\pm$.

Next, we will investigate the Higgs peak, this time around, turning toward the spectral function itself. 
We see in \autoref{fig:resolvent_overview_SC} that the Higgs and phase spectral functions become indistinguishable towards the upper half of the two-particle continuum.
Therefore, we interpret the latter as a background to the Higgs peak itself and compute the difference between both spectral functions.
Using the same methodology as before we obtain that the exponent is $a\approx-0.5$, indicating that the peak behaves as $1 / \sqrt{\omega - \omega_0}$.

@@

Lastly, we investigate the same spectral functions close to the AFM-CDW phase transition, depicted in \autoref{fig:resolvent_overview_AFM}.
We choose $V=1t$ for both lattices and $U=zV \pm 0.1t$, where $z$ is the coordination number.
Again, both SC-related spectral functions are identical in both the CDW and AFM phases.
Within the continuum their weight is shifted towards the upper edge and on the square lattice only, there is a peak above the continuum.
In the AFM phase, both the CDW and AFM spectral functions have a peak close, but yet below, the two-particle continuum.
The AFM peak is at a lower energy than the CDW one.
In the CDW phase, this behavior is reversed, i.e., the CDW peak lies lower.
Using the same analysis as before, we conclude that each of these peaks is a $\delta$ peak.

@@

With the prevalent peaks discussed, we turn towards their behavior as the system approaches the phase transition at $V=0$ for $U=-2.5t$.
First and foremost, we investigate the peak in $\mathcal{A}_\text{Phase}(\omega)$ and $\mathcal{A}_\text{Higgs}(\omega)$ in the CDW phase.
The evolution of its position $\omega_0$ and its weight $w_0$ as $V$ is varied can be viewed double logarithmically in \autoref{fig:sc_in_cdw_behavior}.
For $V \ll 1$, the position grows with the square root of $V$. This behavior then transitions to a purely linear growth for large $V$.
Dividing $\omega_0$ by the gap $\Delta_\text{CDW}$ results in a plateau for large $V$ while leaving the square root behavior at the beginning mostly unchanged.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{plots/phase_peak.pdf}
    \caption{Log-log plot of the real part of $\spectral{Phase}$ in the SC phase at $U=-2.5t$ and $V=-0.1t$ as well as a linear fit to it.
    The upper (lower) panel shows the results for the square (simple cubic) lattice.
    The function behaves as $1/\omega^2$ indicating that the peak in the spectral function is the derivative of a $\delta$ distribution.}
    \label{fig:phase_peak}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=.48\textwidth]{plots/higgs_peak.pdf}
    \caption{Log-log plot of $\spectral{}{Higgs}$ near $\omega = \omega_0 = 2\Delta_\text{SC}$ in the SC phase at $U=-2.5t$ and $V=-0.1t$..
    We additionally plot the difference between $\spectral{}{Higgs}$ and $\spectral{Phase}$ and fit it, as we interpret the latter as a background within the two-particle continuum.
    The upper (lower) panel shows the results for the square (simple cubic) lattice.
    The spectral function behaves as $1 / \sqrt{\omega - \omega_0}$.}
    \label{fig:higgs_peak}
\end{figure}

\begin{figure*}
    \centering
    \includegraphics[width=0.98\textwidth]{plots/peak_fits.pdf}
    \caption{Log-log plot of the real part of various Green's functions, the labels indicate its type and the text in the panels indicate the phase.
        The results for both lattice types are shown but are qualitatively identical.
        The power-law exponent is $a=-1$ in cases, indicating that these peaks are $\delta$ distributions.}
        \label{fig:peak_fits}
\end{figure*}

In this appendix section, we show fits to the real and imaginary parts of various Green's functions in the different phases.
All fits are linear of the type $y = ax + b$ in double logarithmic plots, i.e., $a$ describes the power-law behavior of the functions.
Firstly, the real part of $\greens{}{Phase}$ is shown in \autoref{fig:phase_peak}.
It behaves as $1/\omega^2$ indicating that the peak we find in the spectral function at $\omega=0$ corresponds to the derivative of a $\delta$ distribution.

@@

Lastly, we show all the remaining peaks in \autoref{fig:peak_fits}.
The real part of each one of them behaves as $1/\omega$ in close vicinity to the peak position.
This indicates that the peaks are $\delta$ distributions.
The peaks of $\spectral{AFM}$ and $\spectral{CDW}$ close the the phase transition are identical when swapped, i.e., $\mathcal{A}_\text{AFM} (\omega)$ in the CDW phase is the same as $\mathcal{A}_\text{CDW} (\omega)$ in the AFM phase.

@@

Numerically, we can compute the matrix elements like this
\begin{equation}
    \label{eqn:numerical_example}
    (A_i | A_j) = \frac{1}{N} \sum_{\vk \vl} h(\gamma_i - \hat{\gamma}(\vk)) h(\gamma_j - \hat{\gamma}(\vl)) (A_{\vk} | A_{\vl})\,.
\end{equation}
Here, all expressions are finite numbers and do not require further treatment.
Mutatis mutandis, one obtains a similar expression for $(A_i | [H, A_j])$.
Additionally, note, that the expressions $(A_{\vk} | A_{\vl})$ and $(A_{\vk} | [H, A_{\vl}])$ can be reduced to sums of terms that are proportional to $\gamma_i$ or $\gamma_j$, respectively.
Therefore, it is not necessary to implement any sort of momentum space computations.
For example, consider $A_{\vk} = f_{\vk}^\dagger$, then
\begin{equation}
    ( f_{\vk}^\dagger | f_{\vl}^\dagger ) = \delta_{\vk, \vl} \langle 1 - n_{\vk \uparrow} - n_{\vk \downarrow} \rangle \eqqcolon \delta_{\vk, \vl} \alpha(\vk)
\end{equation}
holds. Inserting this into \eqref{eqn:numerical_example} yields
\begin{align}
    (A_i | A_j) &= \frac{1}{N} \sum_{\vk} h(\gamma_i - \hat{\gamma}(\vk)) h(\gamma_j - \hat{\gamma}(\vk)) \alpha(\vk)\,.
\end{align}
As stated in the main text, see \eqref{eqn:equal_expecs}, $\alpha(\vk)$ depends only on $\hat{\gamma}(\vk)$, hence we can switch from the momentum sum to a $\gamma$ integration
\begin{align}
    (A_i | A_j) &= \int \rho(\gamma) h(\gamma_i - \gamma) h(\gamma_j - \gamma) \alpha(\gamma) \dgamma \nonumber \\
        &\approx \rho(\gamma_i) h(\gamma_j - \gamma_i) \alpha(\gamma_i) \nonumber \\
        &= \frac{\delta_{ij}}{\Delta \gamma} \rho(\gamma_i) \alpha(\gamma_i)\,.
\end{align}
Since $\Delta \gamma$ is a positive constant factor, it does not impact any of our matrix operations, i.e., we can pull it out to the front of the expressions $\mM \to \tilde{m} (\Delta \gamma)^2$ and $\mN \to \tilde{n} (\Delta \gamma)^2$.
Then, we can rewrite \eqref{eqn:green_derivation} as
\begin{equation}
    G_{AB^\dagger} = - (\Delta \gamma)^2 \vec{b}^\dagger (0) \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right] \vec{a}(0)\,.
\end{equation}
To give a practical example, let us compute $\greens{Higgs}$ from \eqref{eqn:resolvent_bases}.
We order our operator set such that all $f_\gamma^{(\dagger)}$ are upfront. 
This requires the vector elements $\vec{a}$ and $\vec{b}$ to be
\begin{equation}
    a_i = b_i = \begin{cases}
        1 & i < 2 N_\gamma \\
        0 & \text{otherwise}
    \end{cases}\,,
\end{equation}
where $N_\gamma$ is the number of discretization points.
Using this, we can expand the relation for the Green's function
\begin{align}
    G_{AB^\dagger} (z) &= -(\Delta \gamma)^2 \sum_{i=0}^{2 N_\gamma} \sum_{j=0}^{2 N_\gamma} \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right]_{ij} \nonumber \\
        &\approx - \int \dgamma \int \dgamma' \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right](\gamma, \gamma')\,.
\end{align}
The first line is an approximation of the integral in the second line, as each matrix element corresponds to a specific value for $\gamma$.
Note, that the sum going up to $2N_\gamma$ rather than $N_\gamma$ merely arises from the fact, that we need to integrate over both $f_\gamma$ and $f_\gamma^\dagger$ to obtain $\greens{Higgs}$.

>>>>

There are a variety of different features. The difference between the two lattices is mostly quantitative.
First and foremost, in the SC phase, see panel (a), there is a sharp peak located at $\omega=0$ in $\spectral{Phase}$ and a singularity in $\spectral{Higgs}$ located at $\omega=2\Delta_\text{SC}$.
We identify them with the well-known Leggett and Higgs modes in superconductors.
The former can be found at zero energy because our model describes a neutral superfluid wherein there is no coupling of the phase of the order parameter to the electromagnetic potential.
Therefore, this phase may be chosen arbitrarily, and changing it requires no energy which ultimately allows a Goldstone mode to exist in the first place. \textcolor{red}{Zitieren!}

@@

This analysis provides that the real part of $\greens{Phase}$ near $\omega=0$ behaves as $1/\omega^2$ indicating that the peak in $\spectral{Phase}$ is the derivative of a $\delta$ distribution.
Note, that it is located at $\omega=0$ of a perfectly asymmetrical bosonic spectral function and therefore must have 0 weight.
This is given here, because $\int \delta'(\omega) \mathrm{d}\omega = 0$.

\textcolor{red}{Ich spekuliere mal...}
The singularity in $\spectral{Higgs}$ behaves as $1/\sqrt{\omega - 2 \Delta_\text{SC}}$.
Previous studies on the dynamics of the order parameter found damped oscillations after a quench.
These have the frequency $\Omega = 2 \Delta_\text{SC}$ and fall off as $1/\sqrt{t}$ \cite{Yuzbashyan06}.
An inverse Fourier transform of $\spectral{Higgs}$ yields precisely the same behavior, thereby reinforcing the validity of our results.

@@

@@

Lastly, we investigate the same spectral functions close to the AFM-CDW phase transition, depicted in \autoref{fig:resolvent_overview_AFM}.
We choose $V=1t$ for both lattices and $U=zV \pm 0.1t$, where $z$ is the coordination number, see panels (a) and (b).
Again, both SC-related spectral functions are identical in both the CDW and AFM phases.
Within the continuum their weight is shifted towards the upper edge and on the square lattice only, there is a peak above the continuum.
In the AFM phase, both $\spectral{AFM}$ and $\spectral{CDW}$ have a peak close, but yet below, the two-particle continuum.
The peak of the former is at a lower energy than the peak of the latter.
In the CDW phase, this behavior is reversed, i.e., the peak in $\spectral{CDW}$ lies lower.
Using the same analysis as before, we conclude that each of these peaks is a $\delta$ peak.
Moving further away from the phase transition, here $U=z(V+0.2t)$ in panels (c) and (d), the upper peak becomes soft as it shifts into the quasiparticle continuum.

@@

With the prevalent peaks discussed, we turn towards their behavior as the system approaches the phase transition at $V=0$ for $U=-2.5t$.
First and foremost, we investigate the peak in $\spectral{Phase}$ and $\spectral{Higgs}$ in the CDW phase.
The evolution of its position $\omega_0$ and its weight $w_0$ as $V$ is varied can be viewed double logarithmically in \autoref{fig:sc_in_cdw_behavior}.
For $V \ll t$, the position grows with the square root of $V$. This behavior then transitions to a purely linear growth for large $V$.
Dividing $\omega_0$ by the gap $\Delta_\text{CDW}$ results in a plateau for large $V$ while leaving the square root behavior at the beginning mostly unchanged.

The peak weights $w_0$ fall off as $1/\sqrt{V}$ for $V \ll t$. 
Afterward, it reaches a shallow minimum and rises slightly to reach a constant value.

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.98\textwidth]{plots/zero_peaks.pdf}
    \caption{The upper panel shows a log-log plot of the real part of $\spectral{Phase}$ in the SC phase at $U=-2.5t$ and $V=-0.1t$ as well as a linear fit to it.
    The lower panel shows the same plot for $\spectral{Higgs}$ in the coexistence phase at $V=0$. The columns show the results for the square and the simple cubic lattice, respectively.
    The functions behave as $1/\omega^2$ indicating that the peaks in the spectral functions are the derivative of a $\delta$ distribution.}
    \label{fig:zero_peaks}
\end{figure*}

\begin{figure}
    \centering
    \includegraphics[width=.48\textwidth]{plots/higgs_peak.pdf}
    \caption{Log-log plot of $\spectral{Higgs}$ near $\omega = \omega_0 = 2\Delta_\text{SC}$ in the SC phase at $U=-2.5t$ and $V=-0.1t$.
    We additionally plot the difference between $\spectral{Higgs}$ and $\spectral{Phase}$ and fit it, as we interpret the latter as a background within the two-particle continuum.
    The upper (lower) panel shows the results for the square (simple cubic) lattice.
    The spectral function behaves as $1 / \sqrt{\omega - \omega_0}$.}
    \label{fig:higgs_peak}
\end{figure}

\begin{figure*}
    \centering
    \includegraphics[width=0.98\textwidth]{plots/peak_fits.pdf}
    \caption{Log-log plot of the real part of various Green's functions, the labels indicate its type and the text in the panels indicate the phase.
        The results for both lattice types are shown but are qualitatively identical.
        The power-law exponent is $a=-1$ in all cases, indicating that these peaks are $\delta$ distributions.}
        \label{fig:peak_fits}
\end{figure*}

In this appendix section, we show fits to the real and imaginary parts of various Green's functions in the different phases.
All fits are linear of the type $y = ax + b$ in double logarithmic plots, i.e., $a$ describes the power-law behavior of the functions.
Firstly, the real part of $\greens{Phase}$ in the SC phase is shown in panel (a) of \autoref{fig:zero_peaks}.
It behaves as $1/\omega^2$ indicating that the peak we find in the spectral function at $\omega=0$ corresponds to the derivative of a $\delta$ distribution.
The same results are obtained for the peak in $\spectral{Higgs}$ and $\spectral{CDW}$ in the coexistence phase.

@@

Lastly, we show all the remaining peaks in \autoref{fig:peak_fits}.
The real part of each one of them behaves as $1/\omega$ in close vicinity to the peak position.
This indicates that the peaks are $\delta$ distributions.
The peaks of $\spectral{AFM}$ and $\spectral{CDW}$ close the the phase transition are identical when swapped, i.e., $\spectral{AFM}$ in the CDW phase is the same as $\spectral{CDW}$ in the AFM phase.

@@

Numerically, we can compute the matrix elements like this
\begin{equation}
    \label{eqn:numerical_example}
    (A_i | A_j) = \frac{1}{N} \sum_{\vk \vl} h(\gamma_i - \hat{\gamma}(\vk)) h(\gamma_j - \hat{\gamma}(\vl)) (A_{\vk} | A_{\vl})\,.
\end{equation}
Here, all expressions are finite numbers and do not require further treatment.
Mutatis mutandis, one obtains a similar expression for $(A_i | [H, A_j])$.
Additionally, note, that the expressions $(A_{\vk} | A_{\vl})$ and $(A_{\vk} | [H, A_{\vl}])$ can be reduced to sums of terms that are proportional to $\gamma_i$ or $\gamma_j$, respectively.
Therefore, it is not necessary to implement any sort of momentum space computations.
For example, consider $A_{\vk} = f_{\vk}^\dagger$, then
\begin{equation}
    ( f_{\vk}^\dagger | f_{\vl}^\dagger ) = \delta_{\vk, \vl}\, \langle 1 - n_{\vk \uparrow} - n_{\vk \downarrow} \rangle \eqqcolon \delta_{\vk, \vl}\, \alpha(\vk)
\end{equation}
holds. Inserting this into \eqref{eqn:numerical_example} yields
\begin{equation}
    (A_i | A_j) = \frac{1}{N} \sum_{\vk} h(\gamma_i - \hat{\gamma}(\vk)) h(\gamma_j - \hat{\gamma}(\vk)) \alpha(\vk)\,.
\end{equation}
As stated in the main text, see \eqref{eqn:equal_expecs}, $\alpha(\vk)$ depends only on $\hat{\gamma}(\vk)$, hence we can switch from the momentum sum to a $\gamma$ integration
\begin{align}
    (A_i | A_j) &= \int \rho(\gamma) h(\gamma_i - \gamma) h(\gamma_j - \gamma) \alpha(\gamma) \dgamma \nonumber \\
        &\approx \rho(\gamma_i) h(\gamma_j - \gamma_i) \alpha(\gamma_i) \nonumber \\
        &= \frac{\delta_{ij}}{\Delta \gamma} \rho(\gamma_i) \alpha(\gamma_i)\,.
\end{align}
Since $\Delta \gamma$ is a positive constant factor, it does not impact any of our matrix operations, i.e., we can pull it out to the front of the expressions $\mM \to \tilde{m} (\Delta \gamma)^2$ and $\mN \to \tilde{n} (\Delta \gamma)^2$.
Then, we can rewrite \eqref{eqn:green_derivation} as
\begin{equation}
    G_{AB^\dagger}(z) = - (\Delta \gamma)^2 \vec{b}^\dagger (0) \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right] \vec{a}(0)\,.
\end{equation}
To give a practical example, let us compute $\greens{Higgs}$ from \eqref{eqn:resolvent_bases}.
We order our operator set such that all $f_\gamma + f_\gamma^{\dagger}$ are upfront. 
This requires the vector elements $\vec{a}$ and $\vec{b}$ to be
\begin{equation}
    a_i = b_i = \begin{cases}
        1 & i < N_\gamma \\
        0 & \text{otherwise}
    \end{cases}\,,
\end{equation}
where $N_\gamma$ is the number of discretization points.
Using this, we can expand the relation for the Green's function
\begin{align}
    \greens{Higgs} (z) &= -(\Delta \gamma)^2 \sum_{i=0}^{N_\gamma} \sum_{j=0}^{N_\gamma} \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right]_{ij} \nonumber \\
        &\approx - \int \dgamma \int \dgamma' \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right](\gamma, \gamma')\,.
\end{align}
The first line is an approximation of the integral in the second line, as each matrix element corresponds to a specific value for $\gamma$.
